signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-16 paper.tex
preliminary results of the analysis

\begin{figure}[ht]\centering
\includegraphics[width=0.55\textwidth]{SP.png}
\caption{The organization of signal peptide.}
\label{fig:sparch}
\end{figure}

@@

The other type of software incorporates into the prediction process theoretical knowledge about the structure of signal peptides. Although these programs do not share innate flaws of ‘black-box’ models, they also demand an improvement. Some of them are based on position matrices or their more abstract variants~\citep{2014zhangprediction, 2004hillerpredisi}. Others (not supported SignalP 3.0, Philius and Phobius) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane}, which reflect regions of signal peptides without realizing limitations of this probabilistic framework. However, the HMMs imply geometric distribution for duration of region length. We replicated the rules for extracting regions' boundaries from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the mentioned assumption is not in concordance with the reality because every region revealed the length distribution other than geometric (Fig. \ref{fig:reglen}). Moreover, the commonly used rigid region scheme (Fig. \ref{fig:sparch}) does not describe extremely long or short signal peptides. Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also unusual structures, but such probabilistic frameworks have not still been implemented.

\begin{figure}[ht]\centering
\includegraphics[width=0.9\textwidth]{reglen.png}
\caption{Distribution of the lengths of signal peptide regions. The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}.
\label{fig:reglen}
\end{figure}

Another feature, shared by majority of the signal peptide predicting software, is orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid. This method of encoding, however, does not preserve relationships between amino acids. The distance between residues is the same regardless of the similarity of their physicochemical properties. Such behaviour is especially undesired in signal peptides models because their regions are defined by specific features of amino acid residues and not by the simple occurrence of specific amino acids. In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}.

@@

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \toprule
Criterion name & Property name \\ 
  \midrule
Size & Size \\ 
   \rowcolor[gray]{0.85}Size & Molecular weight \\ 
  Size & Residue volume \\ 
   \rowcolor[gray]{0.85}Size & Bulkiness \\ 
  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale \\ 
  Hydrophobicity & Hydropathy index \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix \\ 
  Polarity & Polarity \\ 
   \rowcolor[gray]{0.85}Polarity & Mean polarity \\ 
  Frequency in alpha-helices & Signal sequence helical potential \\ 
   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix \\ 
  Frequency in alpha-helices & Relative frequency in alpha-helix \\ 
   \bottomrule
\end{tabular}
\caption{Properties used in clusterization.} 
\label{tab:aaprop}
\end{table}

We selected 13 properties from AAIndex database (aaindexcitation) (see Table~\ref{tab:aaprop}), each attributed to a single criterion. Considering all combinations of single properties for every criterion, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.

@@

@@

You may want to separate results, discussion and conclusion, according to your needs.

Please submit the final pdf file via EasyChair to the GCB'15 program committee by June 30, 2015.

@@

\section*{Acknowledgments}
Thank you for your support!

>>>>

\begin{figure}[ht]\centering
\includegraphics[width=0.55\textwidth]{figures/SP.png}
\caption{The organization of signal peptide.}
\label{fig:sparch}
\end{figure}

@@

The other type of software incorporates into the prediction process theoretical knowledge about the structure of signal peptides. Although these programs do not share innate flaws of ‘black-box’ models, they also demand an improvement. Some of them are based on position matrices or their more abstract variants~\citep{2014zhangprediction, 2004hillerpredisi}. Others (Phobius, Philius and not supported SignalP 3.0) use hidden Markov models (HMMs)~\citep{2004klla, 2008reynoldstransmembrane, 2004bendtsenimproved}, which reflect regions of signal peptides without realizing limitations of this probabilistic framework. However, the HMMs imply geometric distribution for duration of region length. We replicated the rules for extracting regions' boundaries from the first work utilizing HMMs in prediction of signal peptides~\citep{1998nielsenprediction} and found that the mentioned assumption is not in concordance with the reality because every region revealed the length distribution other than geometric (Fig. \ref{fig:reglen}). Moreover, the commonly used rigid region scheme (Fig. \ref{fig:sparch}) does not describe extremely long or short signal peptides. Theoretically, HMMs that describe the atypical signal peptides could be developed to consider also unusual structures, but such probabilistic frameworks have not still been implemented.

\begin{figure}[ht]\centering
\includegraphics[width=0.9\textwidth]{figures/reglen.png}
\caption{A) Distribution of lengths of signal peptides. B) Distribution of the lengths of signal peptide regions. The measure of the length is a number of amino acids in the subsequence. The data was extracted from 2~589 signal peptide sequences derived from UniProt database (see \textbf{Data selection} in \textbf{Methods}).}
\label{fig:reglen}
\end{figure}

Another feature, shared by majority of the signal peptide predicting software, is the orthogonal encoding of amino acids, in which a vector of 20 digits represents every amino acid. This method of encoding, however, does not preserve relationships between amino acids. The distance between residues is the same regardless of the similarity of their physicochemical properties. Such behaviour is especially undesired in signal peptides models because their regions are defined by specific features of amino acid residues and not by the simple occurrence of specific amino acids. In addition to this, such sparse encoding enforces larger data sets, which hinders their management and analysis~\citep{2002linamino}.

@@

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \toprule
Criterion name & Property name \\ 
  \midrule
Size & Size \\ 
   \rowcolor[gray]{0.85}Size & Molecular weight \\ 
  Size & Residue volume~\citep{1973goldsackcontribution} \\ 
   \rowcolor[gray]{0.85}Size & Bulkiness~\citep{1968zimmermanthe} \\ 
  Hydrophobicity & Normalized hydrophobicity scales for alpha-proteins~\citep{1992cidhydrophobicity} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Consensus normalized hydrophobicity scale~\citep{1984eisenbergthreedimensional} \\ 
  Hydrophobicity & Hydropathy index~\citep{1982kytea} \\ 
   \rowcolor[gray]{0.85}Hydrophobicity & Surrounding hydrophobicity in alpha-helix~\citep{1980ponnuswamyhydrophobic} \\ 
  Polarity & Polarity~\citep{1974granthamamino} \\ 
   \rowcolor[gray]{0.85}Polarity & Mean polarity~\citep{1988radzickainfluences} \\ 
  Frequency in alpha-helices & Signal sequence helical potential~\citep{1982argosstructural} \\ 
   \rowcolor[gray]{0.85}Frequency in alpha-helices & Normalized frequency of N-terminal helix \\ 
  Frequency in alpha-helices & Relative frequency in alpha-helix~\citep{1990prabhakaranthe} \\ 
   \bottomrule
\end{tabular}
\caption{Properties used in clusterization.} 
\label{tab:aaprop}
\end{table}

We selected 13 properties from AAIndex database~\citep{2008kawashimaaaindex} (see Table~\ref{tab:aaprop}), each attributed to a single criterion. Considering all combinations of single properties for every criterion, we created 96 possible clusterings of amino acids using Euclidean distance and Ward's method.

@@

\subsection*{n-gram extension}

n-grams (k-mers) are vectors of n characters derived from input sequences. They can be added as the new hidden state to the hidden semi-Markov model. The rigid structure of n-grams may mirror cleavage sites, whose are more conservative than other parts of signal peptide~\citep{2004hillerpredisi}.

Using the biogram software (BIOGRAM CITATION) we extracted n-grams from cleavage sites of signal peptides. XXX n-grams were further used in the analysis

@@

\begin{figure}[ht]\centering
\includegraphics[width=0.9\textwidth]{figures/cvres.png}
\caption{Results of cross-validation for 15 best performing amino acid encodings.}
\label{fig:cvres}
\end{figure}

@@

Despite the presence of signal peptide predicting software, there are no exhaustive studies devoted to the analysis of the signal peptide structure and taxonomical variability. Our research focuses on a transparent model of signal peptide instead of a ‘black-box’ solution.

The proposed algorithm recognizing various signal peptides will reduce costs and speed up searching appropriate artificial signal peptides designed for protein secretion in expensive and time-consuming laboratory experiments. Moreover, the algorithm can be applied in recognition any sequence types on the nucleotide and amino acid levels, represented by small sets. Accessible as a stand-alone and browser-based software, its useful both for researches interested in the analysis of signal peptides.

\section*{Acknowledgments}
PS and PB devised the HSMM model of signal peptide. MB and PS co-wrote the software. MB gathered data and carried the analysis.
MB, PS and PM drafted the manuscript. All authors revised the manuscript and approved the final version.
